A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-05-02 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: adding ensemble space section

>>>>

\section{Ensemble space}

Conceptually, a statistical ensemble is a collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (i.e. a fair coin) or over phase space (i.e. the symplectic manifold represent position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (i.e. a positive semi-definite self-adjoint operator of trace one). An ensemble space is the collection of all possible ensembles defined in a physical theory, together with a basic mathematical structure that all physical theories must possess. The development of a full theory of ensembles goes beyond the scope of this work, and we refer to the appendix and our ongoing work (cite).

For the purpose of this work, we will assume an ensemble space $\Ens$ is a convex subset of a Hausdorff locally convex topological vector space $V$. The topology is the physically natural topology, where each open set represents a verifiable statement. The convex structure represents statistical mixing. For example, $p \ens[a] + \bar{p} \ens[b]$ represent the statistical mixture of two ensembles $\ens[a]$ and $\ens[b]$ with probability $p$ and $\bar{p} = 1 - p$. Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i}^{\infty} p_i \ens_i$ converges in the space or not. We also assume that there is a countable family of affine continuous functions $F_i : \Ens \to \mathbb{R}$ that represent statistical quantities and fully identify each ensemble. That is, $F_i(p \ens[a] + \bar{p} \ens[b]) = p F_i(\ens[a]) + \bar{p} F_i(\ens[b])$ and $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$. For example, we can imagine $F_i$ representing the expectation of all polynomials of position and momentum.\footnote{In this case, the possible probability distributions would be restricted to those that go to zero at infinity on all variables faster than any polynomial. We will leave these technical details out of the main section of the text as to not overwhelm the reader.} Mathematically, these statistical quantities can be used to define semi-norms on the space, giving us a locally convex topology.

Additional, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$. The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensemble that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when they mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

To summarize, the above mathematical structure represents the requirements that ensembles must be experimentally well-defined, allow statistical mixture, can be identified with statistical quantities and provide a well-defined entropy.\footnote{While the full physical justification of the above mathematical structure is beyond the scope of this work, it is important to us that such justification exists and this physically straight forward. This is the essence of what we call Physical Mathematics. We will summarize some of these justifications in the appendix, leaving the full and more updated discussion in our work in progress.} We now turn to the question of how we can represent both the entropy and each ensemble in terms of some measure theoretic structure.
